---
title: "Git, GitHub, and Version Control Quick Tutorial"
author: "Yi Fei Chung"
date: last-modified
format:
  gfm: default

execute:
  eval: false
  echo: true
---

## Table of Contents
[What is Version Control and Git ?](#what-is-version-control-and-git-)\
[How to do Version Control ?](#how-to-do-version-control-)\
[The Pain! Prep work](#the-pain-prep-work)\
[Make the local and the remote talk](#make-the-local-and-the-remote-talk)\
[Core workflow in RStudio](#core-workflow-in-rstudio)\
[Typical RStudio cycle](#typical-rstudio-cycle)\
[Same workflow in VS Code and GitHub Desktop (brief)](#same-workflow-in-vs-code-and-github-desktop-brief)\
[Using GitHub Copilot with Git/GitHub workflows](#using-github-copilot-with-gitgithub-workflows)\
[Quick best practices](#quick-best-practices)\
[Useful coding and GitHub resources](#useful-coding-and-github-resources)\


## What is Version Control and Git ?

- Git is a version control system
- “Track Changes” on steroids
- Ensuring reproducibility
- Ease of working collaboratively
  - With peers
  - With past and future self

## How to do Version Control ?

- Use a hosting service like **GitHub**, Bitbucket, and GitLab
- Think of these as OneDrive, Google Drive or DropBox
- Good even for private solo projects
  - Track changes
  - Revert when you screw up
  - Publication ready
  - Personal website or ShinyApp

## The Pain! Prep work

- Create a [GitHub account](https://github.com/) if you do not already have one.
  - Choose a reasonably professional and recognisable username, since it may appear on your coursework, collaborations, and portfolio.
- *Optional*: Apply for [GitHub Education benefits](https://education.github.com/pack)
  1.  Add and verify your university email address on your GitHub account in **Settings \> Emails**.
  2.  Go to your [GitHub Education benefits settings](https://github.com/settings/education/benefits).
  3.  Under **GitHub Education**, click **Start an application**.
  4.  Complete and submit the application.
- Check whether Git is installed
- Updates (not necessary now)
- Get your local Git to talk to Github *Note*: A **Git repository** (commonly called a "repo") is a digital container or project folder that tracks and stores the complete history of changes made to files over time.

### Check whether Git is installed.

#### macOS

1.  Open **Terminal** by searching for “Terminal” and pressing Return.

2.  Type:

    ```{bash}
    git --version
    ## if you see output similar to `git version 2.39.5 (Apple Git-154)`, Git is installed and ready to use.
    ```

3.  If Git is not installed, macOS may offer to install the Command Line Developer Tools. Follow the prompt to install them.

4.  Once installation is complete, run `git --version` again to check.

#### Windows

1.  Press the Windows key, type **CMD**, and open **Command Prompt**.

2.  Type:

    ```{bash}
    git --version
    ## if you see output similar to `git version 2.50.1.windows.1`, Git is installed and ready to use.
    ```

3.  If Git is not installed, download [Git for Windows](https://gitforwindows.org/).

4.  Follow the installation instructions, then run `git --version` again to check.

## Make the local and the remote talk

There are a few ways to authenticate your local Git with GitHub (i.e., to allow your computer to communicate with your GitHub account). Here I will only cover methods using HTTPS as Personal Access Tokens (PATs). Alternatively, you can also use SSH keys but I found that to be more complicated and less user-friendly for beginners.

- Method 1: Generating a personal access token through GitHub settings
  - At GitHub.com, go to **Settings \> Developer settings \> Personal access tokens \> Tokens (classic) \> Generate new token**.
  - Highly recommend to select `repo`, `workflow`, `gist`, and `user`.
  - **Keep this page open** as you will need to copy the token to your clipboard.
  - **Important**: Never share your personal access token with anyone else.
  - Follow the R code below
- Method 2: Generating a personal access token using the “usethis” package in R
  - Follow the R code below
- Method 3: Using Git Bash / Terminal with SSH keys (not covered here)

```{r}
if(!requireNamespace(c("usethis", "gitcreds"))){
  install.packages(c("usethis", "gitcreds"))
}
library(usethis)
library(gitcreds)

# If you are using Method 1, you can skip this usethis::create_github_token() step
# Generate a personal access token (PAT) for GitHub using the usethis package
usethis::create_github_token()

## Store the personal access token (PAT) 
## This step is the same for both Method 1 and Method 2 after you have generated the token

gitcreds::gitcreds_set()

## Then follow the instructions to paste your personal access token (PAT) into the prompt that appears in RStudio or your terminal. This will securely store your credentials for future Git operations.
```

**GOOD NEWS** : You only need to do this once (or once in a while) per computer!!

## Core workflow in RStudio

1.  **Create a new repository** in GitHub
2.  **Clone the repository** to your local machine using RStudio

```{r}
# Using the usethis package to clone a GitHub repository to your local machine
# Replace "YOUR_USERNAME" and "YOUR_REPO" with your GitHub username
usethis::create_from_github(
  repo_spec = "YOUR_USERNAME/YOUR_REPO",
  destdir = "~/path/to/where/you/want/the/local/repo/",
  fork = FALSE)
```

3.  **Make changes** to your files in RStudio
4.  **Stage and commit** your changes in RStudio
    - *Stage*: Select the files you want to include in your commit
    - *Commit*: Write a concise message describing your changes and commit them to your local repository.
5.  **Pull** any changes from the remote repository (if collaborating)
    - *Pull*: Fetch and merge changes from the remote repository to your local repository. This ensures you have the latest version of the code before pushing your changes.
6.  **Push** your changes to the remote repository on GitHub
    - *Push*: Upload your committed changes from your local repository to the remote repository on GitHub.

### Think of this as you are building a rocket
Your **local repository** is your *base on Earth*, the **remote (GitHub) repository** is your *space station*. You start an almost empty frame on the **remote repository** (your *space station*). Then, you clone it to your **local repository** (your *base on Earth*). You put together the rocket (your code), and when you are ready, you put it on the *launch pad* (**stage**), fuel it up (**commit**), and *launch it* into the space station (**push**). If you are collaborating with others, others may have already launched their rockets into the space station (**push** their changes to the **remote repository**). You need get the latest version of the rocket (**pull**) before you work on the next version of the rocket (your code) again. If not, you may end up with a rocket that does not fit into the space station (**merge conflicts**). 


## Typical RStudio cycle

- Start with a new repository on GitHub.
- Clone it to your local machine. (make sure you have the correct permissions to push to the repository)
- Make change, new files, or delete files.
- Stage relevant files.
- Commit with a concise message.
- Push it to GitHub.
- *If collaborating*, pull any changes from the remote repository before starting your next round of edits.

## Same workflow in VS Code and GitHub Desktop (brief)

- **VS Code**: Use Source Control panel for stage/commit/pull/push; integrated terminal for Git commands.
- **GitHub Desktop**: Use GUI for commit/sync/branching; open repository in editor for coding.
- The logic is still the same: **Start a repo \> clone it \> make changes \> stage \> commit \> push**

## Using GitHub Copilot with Git/GitHub workflows

- Ask Copilot to draft functions, tests, or documentation from comments/prompts.
- Use Copilot Chat to explain code, suggest refactors, or generate commit message drafts.
- **Always review generated code!!!**

## Quick best practices

- Commit often with meaningful messages.
- Keep one logical change per commit.
- Make sure to pull before starting new work.
- Use branches for new features or experiments.
- Resolve conflicts early.

## Useful coding and GitHub resources

- [CBCS Coding Club](https://cbcs-hack.github.io/) by CBCS Coding Club
- [Happy Git and GitHub for the useR](https://happygitwithr.com/) by Jenny Bryan
- [R for Data Science 2nd Edition](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund
- [Advanced R](https://adv-r.hadley.nz/) by Hadley Wickham

## Common commands

### R (RStudio or VS Code)

```{r}
# Check current git version
system("git --version")

# List current git configuration
system("git config --list")

# Create a GitHub personal access token (HTTPS)
usethis::create_github_token()

# Store GitHub personal access token
# (installs or updates credentials used by git)
gitcreds::gitcreds_set()

# Create local project from a GitHub repository
usethis::create_from_github(
  repo_spec = "https://github.com/YOU/YOUR_REPO.git",
  destdir = "~/path/to/where/you/want/the/local/repo/",
  fork = FALSE
)
```

### Terminal (Git Bash)

```{bash}
# Check current git version
git --version

# List current git configuration
git config --list

# Show path to git executable
which git

# Update git on Windows
git update-git-for-windows

# Update git on macOS (Not tested, use with caution)
brew update 
brew upgrade git

# Useful daily workflow
## Get status of your local repository
git status
## Stage files for commit
git add <file>
## Commit staged files with a message
git commit -m "your message"
## Pull changes from the remote repository and rebase
git pull --rebase
## brew update && brew upgrade git
git push
```